% This is the exercise solution document for the homotopy type theory book.

% This file supports two book sizes:
% - Letter size (8.5" x 11")
% - US Trade size (6" x 9")
%
% To activate one or the other, uncomment the appropriate font size in
% the documentclass below, and then one of the two page geometry incantations
%
% NOTE: The 6" x 9" format is only experimental. It will break the
% title page, for example.

\PassOptionsToPackage{table}{xcolor}

% DOCUMENT CLASS
\documentclass[
%
%10pt % for US Trade 6" x 9" book
%
11pt % for Letter size book
]{article}
\usepackage{etex} % We're running out of registers and dimensions, or some such

\newcounter{chapter}            % So that macros.tex doesn't choke

% PAGE GEOMETRY
%
% Uncomment one of these

% We make the page 40pt taller than the standard LaTeX book.

% OPTION 1: Letter
\usepackage[papersize={8.5in,11in},
            twoside,
            includehead,
            top=1in,
            bottom=1in,
            inner=0.75in,
            outer=1.0in,
            bindingoffset=0.35in]{geometry}

% OPTION 2: US Trade
% \usepackage[papersize={6in,9in},
%             twoside,
%             includehead,
%             top=0.75in,
%             bottom=0.75in,
%             inner=0.5in,
%             outer=0.75in,
%             bindingoffset=0.35in]{geometry}

% HYPERLINKING AND PDF METADATA

\usepackage[pagebackref,
            colorlinks,
            citecolor=darkgreen,
            linkcolor=darkgreen,
            unicode,
            pdfauthor={Univalent Foundations Program},
            pdftitle={Homotopy Type Theory: Univalent Foundations of Mathematics},
            pdfsubject={Mathematics},
            pdfkeywords={type theory, homotopy theory, univalence axiom}]{hyperref}

% OTHER PACKAGES

% Use this package and stick \layout somewhere in the text to see
% page margins, text size and width etc. Useful for debugging page format.
%\usepackage{layout}

%%% Because Germans have umlauts and Slavs have even stranger ways of mangling letters
\usepackage[utf8]{inputenc}

%%% For table {tab:theorems}
\usepackage{pifont}

%%% Multi-Columns for long lists of names
\usepackage{multicol}

%%% Set the fonts
\usepackage{mathpazo}
\usepackage[scaled=0.95]{helvet}
\usepackage{courier}
\linespread{1.05} % Palatino looks better with this

\usepackage{graphicx}
\usepackage{comment}

\usepackage{wallpaper} % For the background image on the cover page

\usepackage{fancyhdr} % To set headers and footers

\usepackage{nextpage} % So we can jump to odd-numbered pages

\usepackage{amssymb,amsmath,amsthm,stmaryrd,mathrsfs,wasysym}
\usepackage{enumitem,mathtools,xspace}
\usepackage{xcolor} % For colored cells in tables we need \cellcolor
\usepackage{booktabs} % For nice tables
\usepackage{array} % For nice tables
\usepackage{supertabular} % For index of symbols
\definecolor{darkgreen}{rgb}{0,0.45,0}
\usepackage{aliascnt}
\usepackage[capitalize]{cleveref}
\usepackage[all,2cell]{xy}
\UseAllTwocells
\usepackage{natbib}
\usepackage{braket} % used for \setof{ ... } macro
\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing}

\usepackage{etoolbox}           % hacking commands for TOC

\usepackage{mathpartir}         % for formal.tex appendix, section 3

\usepackage[numbered]{bookmark} % add chapter/section numbers to the toc in the pdf metadata

\input{macros}

%%%% Indexing
\usepackage{makeidx}
\makeindex

%%%% Header and footers
\pagestyle{fancyplain}
\setlength{\headheight}{15pt}
\renewcommand{\sectionmark}[1]{\markright{\textsc{\thesection\ #1}}}

\lhead[\fancyplain{}{{\thepage}}]%
      {\fancyplain{}{\nouppercase{\rightmark}}}
\rhead[\fancyplain{}{\nouppercase{\leftmark}}]%
      {\fancyplain{}{\thepage}}
\cfoot{\textsc{\scriptsize \textcolor{gray}{[Draft of \today]}}}
\lfoot[]{}
\rfoot[]{}

%%%% Chapter & part style
\usepackage{titlesec}
\titleformat{\part}[display]{\fontsize{40}{40}\fontseries{m}\fontshape{sc}\selectfont}{\hfil\partname\ \Roman{part}}{20pt}{\fontsize{60}{60}\fontseries{b}\fontshape{sc}\selectfont\hfil}
\titleformat{\chapter}[display]{\fontsize{23}{25}\fontseries{m}\fontshape{it}\selectfont}{\chaptertitlename\ \thechapter}{20pt}{\fontsize{35}{35}\fontseries{b}\fontshape{n}\selectfont}

%\input{main.labels}

\title{Solutions to selected exercises}

\begin{document}
\maketitle

\section*{Exercises from \autoref{cha:typetheory}}

\subsection*{Solution to \autoref{ex:composition}}

We of course know what composition of functions $f : A \to B$ and $g : B \to C$ should be, but let us see how we might derive it by considering available forms of construction. We want a term
\[ g \circ f \defeq (\Box : A \to C), \]
where $\Box : A \to C$ indicates that in place of $\Box$ we would like to put something of type $A \to C$.
Since we are defining a function whose domain is $A$, we expect it to be of the form
\[ g \circ f \defeq \lam{x:A} (\Box : C), \]
so now we are looking for something of type $C$, with $x$, $f$ and $g$ available. Of these $g$ looks most promising as it lands in $C$:
\[ g \circ f \defeq \lam{x:A} g (\Box : B). \]
Now we repeat the same trick with $f$ to get
\[ g \circ f \defeq \lam{x:A} g(f(\Box : A)). \]
Inside the abstraction $x$ is available and has the type we need, so we define
\begin{equation}
  \label{eq:composdef}
   g \circ f \defeq \lam{x:A} g(f(x)) : C
\end{equation}
%
This baby example demonstrates how one often works with a proof assistant: look at what
you need and what is available, and try to make some progress.

Now, suppose given also $h : C \to D$. We have, according to \autoref{eq:composdef},
%
\begin{align*}
  h \circ (g \circ f) &\jdeq \lamu{x:A} h ((\lam{y:A} g(f(y))) x)\\
                      &\jdeq \lamu{x:A} h(g(f(x))),
\end{align*}
%
and
%
\begin{align*}
  (h \circ  g) \circ f & \jdeq \lamu{x:A} (\lam{y:A} h(g(y))) (f(x))\\
                       & \jdeq \lamu{x:A} h(g(f(x))).
\end{align*}
%
They are equal, which establishes associativity of composition.

\subsection*{Solution to \autoref{ex:pr-to-rec}}

If we suppose given only $\fst : A \times B \to A$ and $\snd : A \times B \to B$ satisfying $\fst(\tup{a}{b}) \jdeq a$ and $\snd(\tup{a}{b})\jdeq b$, we can define $\rec{A\times B}'$ by
\[ \rec{A\times B}'(C,g,x) \defeq g (\fst x) (\snd x). \]
We can now verify, given $C:\UU$, $g:A\to B \to C$ and $(a,b):A\times B$,
\begin{align*}
  \rec{A\times B}'(C,g,(a,b)) &\jdeq g (\fst (a,b)) (\snd (a,b))\\
                              &\jdeq g (a) (b).
\end{align*}
%
For $\Sigma$-types we replace $A \times B$ above with $\sm{a:A} B(a)$, but otherwise
everything else stays the same:
\[ \rec{\sm{x:A} B(x)}'(C,g,x) \defeq g (\fst x) (\snd x). \]

\subsection*{Solution to \autoref{ex:pr-to-ind}}

Quite naturally, we form
\[ \ind{A\times B}''(C,g,x) \defeq g (\fst x) (\snd x),\]
of type
\[ \prd{C:A\times B \to \UU}\Parens{\prd{y:A}\prd{z:B} C((y,z))} \to 
   \prd{x : A \times B} C ((\fst x,\snd x)). \]
This is not quite what we need because $\ind{A\times B}$ has the type
\[ \prd{C:A\times B \to \UU}\Parens{\prd{y:A}\prd{z:B} C((y,z))} \to
   \prd{x : A \times B} C (x). \]
%
Recall that we have the propositional uniqueness principle
%
\[ \uppt: \prd{x : A \times B} ((\fst x,\snd x)=_{A\times B} x), \]
%
satisfying $\uppt{(a,b)} = \refl{(a,b)}$.
We can transport along $\uppt(x)$ to get from $C((\fst x, \snd x))$ to $C(x)$:
%
\[ \ind{A\times B}'(C,g,x) \defeq 
    \transfib{C}{\uppt(x)}{\ind{A \times B}''(C, g, x)}.
\]
%
It remains to verify that $\ind{A \times B}'(C, g, x)$ behaves as expected:
%
\begin{align*}
  \ind{A \times B}'(C,g,(a,b))
  &\jdeq \transfib{C}{\uppt((a,b))}{g(\fst (a,b))(\snd(a,b))} \\
  &\jdeq \transfib{C}{\uppt((a,b))}{g(a)(b)} \\
  &\jdeq \transfib{C}{\refl{(a,b)}}{(g(a)(b)} \\
  &\jdeq g(a)(b).
\end{align*}
%
Now for $\Sigma$-types the exact same expressions work as well, except that the types change.

\section*{Exercises from \autoref{cha:logic}}

\subsection*{Solution to \autoref{ex:decidable-choice}}

The hypotheses imply that
\[ \Parens{\sm{n:\nat}P(n)} \to \sm{n:\nat}\Parens{P(n) \times \prd{m:\nat} \big((m<n) \to \neg P(m)\big)}. \]
In words, given $n$ such that $P(n)$, we can find the least such $n$: we test every $m<n$ in turn, using decidability to do a case analysis, until we find the first one that satisfies $P(m)$.
However, the right-hand side of the above implication is a mere proposition: if both $n$ and $n'$ are least numbers satisfying~$P$ then they must be equal.
Therefore, we also have
\[ \Brck{\sm{n:\nat}P(n)} \to \sm{n:\nat}\Parens{P(n) \times \prd{m:\nat} \big((m<n) \to \neg P(m)\big)} \]
from which the claim follows.

\section*{Exercises from \autoref{cha:equivalences}}

\subsection*{Solution to \autoref{ex:symmetric-equiv}}

First note that for any type $A$ we have $\eqv{\iscontr(A)}{A\times \iscontr(A)}$.
Thus
\begin{align*}
  \prd{a:A} \iscontr\Parens{\sm{b:B} R(a,b)}
  &\eqvsym \prd{a:A} \Parens{\sm{b:B} R(a,b)} \times  \iscontr\Parens{\sm{b:B} R(a,b)}\\
  &\eqvsym \Parens{\prd{a:A}\sm{b:B} R(a,b)} \times  \Parens{\prd{a:A}\iscontr\Parens{\sm{b:B} R(a,b)}}\\
  &\eqvsym \Parens{\sm{f:A\to B} \prd{a:A} R(a,f(a))} \times  \Parens{\prd{a:A}\iscontr\Parens{\sm{b:B} R(a,b)}}
\end{align*}
using \autoref{thm:ttac} at the last step.
So the type given in the exercise is equivalent to
\begin{equation*}
\sm{f:A\to B}{R:A\to B\to \type}
\Parens{\prd{a:A} R(a,f(a))}\times
\Parens{\prd{a:A} \iscontr\Parens{\sm{b:B} R(a,b)}} \times
\Parens{\prd{b:B} \iscontr\Parens{\sm{a:A} R(a,b)}}.
\end{equation*}
It will therefore suffice to show that for any $f:A\to B$, the type
\begin{equation}\label{eq:symmetric-isequiv}
\sm{R:A\to B\to \type}
\Parens{\prd{a:A} R(a,f(a))}\times
\Parens{\prd{a:A} \iscontr\Parens{\sm{b:B} R(a,b)}} \times
\Parens{\prd{b:B} \iscontr\Parens{\sm{a:A} R(a,b)}}.
\end{equation}
is equivalent to $\isequiv(f)$, or equivalently that it satisfies the three desiderata of $\isequiv(f)$.

Firstly, suppose $f$ has a quasi-inverse $g$, and define $R(a,b) \defeq (f(a)=b)$.
For any $a$ we have $\iscontr(\sm{b:B} R(a,b))$ by \autoref{thm:contr-paths}, and in particular we have $R(a,f(a))$.
On the other hand, by \autoref{thm:paths-respects-equiv} we have $\eqv{R(a,b)}{(gf(a) = g(b))}$, which is equivalent to $a = g(b)$, so \autoref{thm:contr-paths} also implies that $\iscontr(\sm{a:A} R(a,b))$ for any $b:B$.

Secondly, suppose~\eqref{eq:symmetric-isequiv} is inhabited, i.e.\  we have $R:A\to B\to \type$ and witnesses $r:\prd{a:A} R(a,f(a))$ and $c:\prd{a:A} \iscontr(\sm{b:B} R(a,b))$ and $d:\prd{b:B} \iscontr(\sm{a:A} R(a,b))$.
Let $g(b) \defeq \proj1 (\proj1(d(b)))$, yielding $g:B\to A$; thus we have $R(g(b),b)$ for any $b:B$.
Then for any $a_0:A$ we have $R(a_0,f(a_0))$ and $R(g(f(a_0)),f(a_0))$; but $\sm{a:A} R(a,f(a_0))$ is contractible, so $a_0 = g(f(a_0))$.
Similarly, $b_0 = f(g(b_0))$ for any $b_0:B$, so $g$ is a quasi-inverse to $f$.

Finally, we must show that~\eqref{eq:symmetric-isequiv} is a mere proposition.
Since $\prd{b:B} \iscontr\Parens{\sm{a:A} R(a,b)}$ is a mere proposition by \cref{thm:isprop-forall,thm:isprop-iscontr}, by \autoref{thm:path-subset} we may ignore it and consider only the remainder:
\begin{equation*}
\sm{R:A\to B\to \type}
\Parens{\prd{a:A} R(a,f(a))}\times
\Parens{\prd{a:A} \iscontr\Parens{\sm{b:B} R(a,b)}}.
\end{equation*}
Using \autoref{thm:ttac} again, this is equivalent to
\begin{equation*}
\prd{a:A}\sm{R:B\to \type} R(fa)\times \iscontr\Parens{\sm{b:B} R(b)}.
\end{equation*}
Thus it will suffice to show that for any $b_0:B$, the type
\[ \sm{R:B\to \type} R(b_0)\times \iscontr\Parens{\sm{b:B} R(b)} \]
is a mere proposition.
But in fact, this type is contractible; its center of contraction consists of $\lam{b}(b_0=b)$ and $\refl{b_0}$ and \autoref{thm:contr-paths}, and the contracting homotopy arises from \autoref{thm:identity-systems}\ref{item:identity-systems4}$\Rightarrow$\ref{item:identity-systems3} (together with univalence and function extensionality).


\section*{Exercises from \autoref{cha:hlevels}}

\subsection*{Solution to \autoref{ex:ntype-from-nconn-const}}

Let $\modal$ be a modality; it remains to show that if all maps into $B$ out of a $\modal$-connected type are constant, then $B$ is $\modal$-modal.
We show under that hypothesis that $\eta:B\to \modal B$ is an equivalence, by showing that its fibers are contractible.

Since $\eta$ is always $\modal$-connected, its fibers are $\modal$-connected, and thus the inclusion of $\hfib{\eta}{z}$ into $B$ is constant.
Thus, there is an $a:B$ such that for any $b:B$ and $p:\eta b=z$ we have $q(b,p):a=b$.
Then $\apfunc{\eta}(q(b,p))\ct p : \eta a=z$ for any $(b,p):\hfib\eta z$, so we have a map $\hfib\eta z \to (\eta a=z)$.
But $\modal B$ is $\modal$-modal, hence so is $\eta a = z$.
Thus, by \autoref{thm:nconn-to-ntype-const}, this map is constant; hence we have $r:\eta a=z$ such that $\apfunc{\eta}(q(b,p))\ct p = r$ for all $(b,p)$.
But using the characterization of paths in fibers, this means exactly that $(b,p)=(a,r)$ for all $(b,p)$; hence $\hfib\eta z$ is contractible with center $(a,r)$.

\subsection*{Solution to \autoref{ex:connectivity-inductively}}

If $A$ is $n$-connected, then since $\trunc{-1}A = \trunc{-1}{\trunc{n}A}$, also $A$ is $(-1)$-connected.
And since $\trunc{n-1}{\id[A]ab} = (\id[\trunc n A]{\tproj na}{\tproj nb})$ by \autoref{thm:path-truncation} and the path spaces of a contractible type are contractible, each $\id[A]ab$ is $(n-1)$-connected.

Conversely, suppose $A$ is $(-1)$-connected and all its path spaces are $(n-1)$-connected.
Firstly, we claim $\trunc nA$ is a mere proposition, i.e.\ that for all $x,y:\trunc nA$, the type $x=y$ is contractible.
Since contractibility of $x=y$ is a mere proposition, it suffices to assume that $x$ and $y$ are of the form $\tproj na$ and $\tproj nb$ respectively.
But $\id[\trunc n A]{\tproj na}{\tproj nb}$ is contractible by \autoref{thm:path-truncation} and the assumption.
Thus, $\trunc nA$ is a mere proposition.
Since it is also $(-1)$-connected by assumption, it is therefore contractible, so $A$ is $n$-connected.


\subsection*{Solution to \autoref{ex:lemnm}}

Evidently $\mathsf{LEM}$ is the same as $\mathsf{LEM}_{-1,-1}$, so for the first part it suffices to assume $\mathsf{LEM}$ and prove $\mathsf{LEM}_{n,-1}$ and $\mathsf{LEM}_{-1,m}$ for all $n,m$.
For the latter, note that if $A$ is a mere proposition, then by \autoref{ex:lem-mereprop} so is $A+\neg A$, and thus $\trunc m{A+\neg A}= A+\neg A$ for any $m\ge -1$.
For the former, note that assuming LEM, by \autoref{ex:lem-brck} we have $\trunc{-1}{B} = \neg\neg B$ for any $B$, while $\neg\neg(A+\neg A)$ is always true for any type $A$ (\autoref{ex:not-not-lem}).

For the second part, it suffices to derive a contradiction from $\mathsf{LEM}_{0,0}$; but the proof of \autoref{thm:not-lem} already uses an $A$ that is a set (namely $\bool$).

\subsection*{Solution to \autoref{ex:acconn}}

Suppose the $(-1)$-connected AC, and by induction suppose also the $n$-connected AC.
Let $X$ be a set and $Y:X\to \type$ a family of $(n+1)$-connected types.
By \autoref{ex:connectivity-inductively}, to show that $\prd{x:X} Y(x)$ is $(n+1)$-connected, it suffices to show that it is $(-1)$-connected and that all its path-types are $n$-connected.
But also by \autoref{ex:connectivity-inductively}, each $Y(x)$ is $(-1)$-connected and all its path types are $n$-connected.
Applying function extensionality to characterize the path types of $\prd{x:X} Y(x)$, the claim follows from the $(-1)$-connected and $n$-connected axioms of choice.

\section*{Exercises from \autoref{cha:category-theory}}

\subsection*{Solution to \autoref{ex:stack}}

Define $K$ to be the precategory with $K_0 \defeq Y$ and $\hom_K(y_1,y_2) \defeq (p(y_1)=p(y_2))$.
Then $\mathrm{Desc}(A,p)\defeq A^K$ is a good definition.
Moreover, the obvious functor $K\to X$ (where $X$ denotes the discrete category on itself) is a weak equivalence, so \autoref{ct:esofull-precomp-ff,ct:cat-weq-eq} yield the second and third parts.
Finally, $K$ is a strict category, so if it is a stack, then $p$ has a section, while conversely if $p$ has a section then $K\to X$ is a (strong) equivalence.

\section*{Exercises from \autoref{cha:set-math}}

\subsection*{Solution to \autoref{ex:prop-ord}}

Define $A<B$ iff $B \land \neg A$.

\subsection*{Solution to \autoref{ex:ninf-ord}}

Define $a<b$ iff $\exis{n:\nat} (b_n < a_n)$.

\end{document}
